Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-17 paper/paper.tex
more edits to Results

\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
\usepackage[export]{adjustbox}

\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references

@@

The SRA is designed as a permanent archive of deep sequencing data.
The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.

\begin{figure}[]
\centering
\fbox{\includegraphics[trim={0.8in 0.8in 1.1in 1in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
\caption{Example of the process to delete SRA data.
The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff.
The full e-mails are at \url{https://usrtk.org/tag/pangolin-papers/}.
}
\label{fig:pangolin_emails}
\end{figure}

@@

\subsection{Recovery of deleted sequencing data from the Google cloud} 
As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
For instance, as of June 3, 2020, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.

@@

Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
This strategy was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.

@@

\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
The approach of \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by a single rather than nested PCR.
I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.

@@

\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
To contextualize the viral sequences recovered from the deleted project described above, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.

@@

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

@@

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which is reported to have been collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as implied by phylogenetic rooting.
\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the bat outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
This explanation can be discounted, because although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other unknown market being the source of viruses that lack the T8782C and C28144T mutations.
However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows this idea can also be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.

@@

Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any sequences from the Huanan Seafood Market across the region in which all were fully sequenced (Figure~\ref{fig:deltadist_jitter}).
To place this observation in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees described in the previous section (Figure~\ref{fig:tree_with_deleted}).

It is immediately apparent that the discrepancy between outgroup rooting and epidemiological evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong (Figure~\ref{fig:tree_with_deleted}).
The middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
In this scenario, the progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a putative progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
Note that although three deleted early epidemic sequences are compatible with this proCoV2 progenitor, the defining C18060T mutation is not covered in the sequenced region---so further sequencing of these samples would be informative.
The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.

We can also qualitatively evaluate the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the sample principle employed by \citet{worobey2020emergence} to help evaluate scenarios for the emergence of SARS-CoV-2 in Europe and North America: namely that during an exponentially growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
This principle is especially likely to hold for the scenarios represented in Figure~\ref{fig:tree_with_deleted}, where the are multiple observed representatives of each putative progenitor which each have the potential to yield descendants with new mutations.
Using this qualitative idea, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
I acknowledge that these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in more qualitative reasoning when there are valid concerns about the nature of the underlying data.

>>>>

\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
\usepackage[export]{adjustbox}
\usepackage{xurl}

%\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
%\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references

@@

The SRA is designed as a permanent archive of deep sequencing data.
The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} requesting deletion of two sequencing runs.
Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.

\begin{figure}[]
\centering
\fbox{\includegraphics[trim={0.8in 0.8in 1.1in 1in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
\caption{Example of the process to delete SRA data.
The image shows e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff excerpted from \url{https://usrtk.org/tag/pangolin-papers/}.
}
\label{fig:pangolin_emails}
\end{figure}

@@

\subsection{Recovery of deleted sequencing data from the Google cloud} 
As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
For instance, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned the message ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.

@@

Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
This strategy was successful; for instance, as of June 3, 2021, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.

@@

\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
The approach of \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.

@@

\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
To contextualize the viral sequences recovered from the deleted project, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.

@@

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

@@

This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as implied by phylogenetic rooting.
\citet{pipes2021assessing} discuss the conundrum in detail, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

Before moving to the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
But although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other market being the source of viruses that lack the T8782C and C28144T mutations.
However, inspection of Figure~\ref{fig:tree_RaTG13} shows that clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.

@@

Repeating the analysis of the previous section with these new sequences shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology analyses in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.

Furthermore, it is immediately apparent that the discrepancy between outgroup rooting and the evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong.
The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
This putative progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
However, analysis of this rooting is limited by the fact that the defining C18060T mutation is in a region not covered in the deleted sequences.
The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.

We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.
